\vspace{-0.2in} Parallelization of Fast $\ell_1$-Minimization for Face Recognition

### 2011-06-05 main.tex
Fixed some awkward sentences, make fig. 4 bigger.

%\title{Accelerating $\ell_1$-Minimization Using Many-Core CPUs/GPUs \\ and Application to Face Recognition
%\title{Efficient Parallelization of Sparse Representation for Face Recognition}
%\thanks{Corresponding author: . This work was partially supported by ARO MURI W911NF-06-1-0076.}}
%\title{Many-Core Parallelization of $\ell_1$-Minimization for Face Recognition
\title{Parallelization of Fast $\ell_1$-Minimization for Face Recognition}
%\title{Parallelization of Fast $\ell_1$-Minimization Towards Real-Time Face Recognition}

@@

\begin{abstract} Recently a family of promising face recognition algorithms
based on sparse representation and $\ell_1$-minimization ($\ell_1$-min) have been
developed.  These algorithms have not yet seen commercial application, 
largely due to higher computational cost compared to other traditional
algorithms. This paper studies techniques for leveraging the 
massive parallelism available in GPU and CPU hardware to accelerate
$\ell_1$-min based on augmented Lagrangian method (ALM) solvers. 
Due to the low arithmetic intensity of ALM solvers, implementations on both
architectures are primarily limited by memory bandwidth,  
with large problems running faster on the GPU and small problems running faster
on the CPU.  
On both architectures, the proposed implementations significantly outperform
naive library-based implementations. 
At the problem sizes most important for $\ell_1$-min based
face recognition, both architectures achieve comparable performance.
The source code of the
algorithms will be made available for peer evaluation.  \end{abstract}
\vspace{-0.1in}

@@

In this paper, we study parallelization of $\ell$-min on many-core CPU and GPU
architectures. Although $\ell_1$-min \eqref{eq:l1min} is a convex
program, conventional algorithms such as interior-point methods
\cite{ChenS2001-SIAM,TibshiraniR1996} scale poorly
for applications such as face recognition. Recently, a number of
accelerated algorithms have been proposed that explicitly take advantage of
the special structure of $\ell_1$-min problems
\cite{LorisI2009,YangA2010-ICIP}. We investigate parallelization of a
state-of-the-art $\ell_1$-min solution based on the classical framework of
\emph{augmented Lagrangian methods} (ALM) \cite{BertsekasD2003,YangA2010-ICIP}.

@@

The $ell_1$ solvers in the original system prototype were based 
on standard interior-point methods \cite{WagnerA2009-CVPR}. Their speed has been improved by ALM in a more recent paper \cite{WagnerA2011-PAMI}. However, due the high per-class cost
of the alignment step, the recognition system still fails to achieve real-time
recognition against datasets of hundreds or thousands of subjects.  In this paper,
in addition to accelerating the generic
$\ell_1$-min objectives \eqref{eq:l1min} and \eqref{eq:l1min_denoise}, we 
also discuss how to efficiently accelerate the iterative face alignment step
\eqref{eq:l1min_alignment} on multi-core CPUs and GPUs.

@@

Finally, we present extensive benchmark to compare the performance of the
generic ALM $\ell_1$-min algorithm and the corresponding face recognition
routines on massively parallel CPUs and GPUs.  To this end, a face
recognition system for security and access control applications has been implemented, 
and is diagrammed in Figure \ref{fig:pipeline}. This pipeline follows the work of \cite{WagnerA2011-PAMI}, but
the fast implementations presented in this paper are novel.
Since the parallelization of any algorithm
is highly dependent on the target hardware, we will give a brief overview of
the architecture of our test system in detail in Section
\ref{sec:concurrency}.
Finally, note that 
while this paper focuses on the parallelism available in a single (shared memory) machine,
applications targeting thousands of subjects or more, will likely require computation be
distributed over a cluster or network of computers, and this is not addressed here.
The source code of our implementations, which target NVIDIA Fermi GPUs
as well as Intel x86-64 CPUs, will be made available for peer evaluation.
\begin{figure}
\centering
{\tiny \input{figures/pipeline_simplified.tex}}
\caption{The face recognition pipeline.\vspace{-0.1in}}
\label{fig:pipeline}
\end{figure}

@@

One of the drawbacks of most interior-point methods for $\ell_1$-min (including
BP, LASSO, and TNIPM) is that they require the solution sequence to follow an
interior path via gradient descent or conjugate gradient methods, which are computationally expensive.
To mitigate these issues, an approach called \emph{Homotopy} has been recently studied to accelerate the
speed of $\ell_1$-min. Homotopy methods for recovering sparse signals were
first studied in the context of LASSO \cite{OsborneM2000}, which inspired a
solution to the \emph{forward stagewise linear regression} problem called LARS
\cite{EfronB2004}, and eventually led to the corresponding Homotopy algorithms
for BP in \cite{MalioutovD2005,DonohoD2006}.

Although Homotopy can be shown to exactly estimate BP when the solution is
sufficiently sparse \cite{DonohoD2006}, the algorithm still involves
computationally expensive operations such as matrix-matrix multiplication and
linear least-squares problems with varying $A$
matrices.
%\footnote{During the face alignment algorithm presented later, The $A$ matrix changes between alignment iterations but not within the L1 solver}  
In Section \ref{sec:ALM}, we contend that ALM is a better choice for
implementation on many-core CPUs and GPUs. The ALM algorithm belongs to a
category of approximate $\ell_1$-min solutions called \emph{iterative
shrinkage-thresholding} (IST) methods \cite{WrightS2008,BeckA2009}.  IST
algorithms mainly involve elementary operations such as vector algebra and
matrix-vector multiplication. Therefore, when the dimension of the problem
becomes high, IST-type algorithms are particularly suitable for hardware
systems with a high degree of concurrency. In \cite{YangA2010-ICIP}, the
authors showed that ALM is able to significantly improve the solver speed,
while achieving estimation accuracy competitive with exact $\ell_1$-min
solutions such as BP and Homotopy methods, and other IST-type
algorithms. Therefore, we choose ALM as the core algorithm for
implementation of $\ell_1$-min in the parallel face recognition pipeline.

@@

In practice, the optimal values for the triplet $(\xx^*, \yy^*, \mu)$ are all
unknown. Furthermore, it has been observed that solving
\eqref{eq:optimal-ALM} with a large initial value of $\mu$ tends to lead to
slower convergence speed \cite{WrightS2008,YangA2010-ICIP}. In
\cite{BertsekasD2003,YangJ2009}, an alternating procedure has been shown to
iteratively update $\xx$ and $\yy$:
\begin{equation}
\left \{
\begin{array}{lll}
\xx_{k+1} & = & \arg\min_{\xx} \, L_{\mu_k} (\xx,\yy_k)\\
\yy_{k+1} & = & \yy_k + \mu_k (\bb - A\xx_{k+1}) \\
\end{array}
\right . ,
\label{eq:alm}
\end{equation}
where $\mu_{k}\rightarrow \infty$ is monotonically increasing.
The iteration terminates when the estimates $(\xx_k, \yy_k)$ converge.

@@

\subsection{GPU Hardware Concurrency} \vspace{-0.05in}
The main defining characteristics of contemporary multi-core GPU architectures
are that they have two (much wider) levels of concurrency, relatively small amounts of
 cache, and relatively low clock speeds.  Whereas most of the transistors on
a typical CPU are dedicated either to cache or to hardware that enables higher
clock speeds (such as branch prediction, out-of-order instruction execution, etc...),
most of the transistors on a GPU are dedicated to arithmetic logic units.
For our GPU implementations we target NVIDIA Fermi GPU architecture (\eg, the GTX 480
GPU used in this paper).  An explanation of the GPU programming model (CUDA) at a useful level of completeness would take more space than is available here, so we will instead frame our discussion in terms of 
hardware capabilities. 
%\footnote{A discussion of the CUDA programming model can be found in the NVIDIA C for CUDA Programming Guide}
CUDA programmable GPUs are comprised of several
\emph{streaming multiprocessors} (SMP), each of which is roughly analogous to a CPU core.
For Fermi architecture GPU's, there are up to 16 SMP's, and each SMP is capable of executing up to 64 single precision 
floating point operations
concurrently. \footnote{In CUDA terms, the warp width is 32, and the floating pipeline can
issue up to two warps simultaneously}

@@

While the programming model for
the SIMD units of each SMP is somewhat more flexible than on the CPU,
leveraging the flexibility typically comes at the cost of reduced concurrency.
\footnote{In CUDA terms, code branches that cause warp divergence usually
result in serialization of the different groups of threads}  Therefore, for
the purposes of comparing hardware architectures, the SIMD units on the GPU are
roughly analogous to the CPU SIMD units. Thus, in summary, the GPU hardware 
provides two levels of concurrency: {\em SMP-level}
concurrency and {\em thread-level} concurrency.  Note that the GPU provides more
concurrency than a CPU at both levels (14 SMP's vs. 8 cores) and (64 wide vs. 4 wide SIMD
units).

@@

Face alignment \eqref{eq:l1min_alignment} estimates an image transformation
$\tau$ that rectifies the query image $\tilde{\bb}$ with possible pose
variation w.r.t. each training class $A_i$, which leads to the minimal sparsity
in error $\ee$ after the alignment. Note that directly solving
\eqref{eq:l1min_alignment} is impossible since it is a non-convex problem
and may exhibit local minima.
However, given a good initial guess of the
transformation $\tau$ (\eg, provided by an accurate face detector), the optimal
solution for $\tau$ can be sought iteratively by linearization at each $j$th
step:
\begin{equation}
\min_{\xx, \ee, \Delta \tau_j}\|\ee\|_1\quad \subj\quad \tilde\bb\circ\tau_j +  J_j\Delta \tau_j = A_i\xx + \ee,
\end{equation}
where $J_j = \nabla_{\tau_j}(\tilde\bb\circ\tau_j)$ is the Jacobian, and
$\Delta \tau$ is a step update to $\tau$. Denote $\bb_j =
\tilde\bb\circ\tau_j$, $B_j = [A_i, -J_j]$ and $\ww^T = [\xx^T, \Delta
\tau^T]$, then the update $\Delta \tau$ can be computed by solving the
following linear program:
\begin{equation}
\min_{\ww, \ee}\|\ee\|_1\quad \subj\quad \bb_j = B_j\ww + \ee.
\label{eq:alignment-linearization}
\end{equation}
%If the warping $\tau$ is constrained to be a similarity transform,
%it can be paramaterized in such a way that the computation of $\frac{\partial \bb}{\partial \tau_p}$
%reduces to several linear vector operations involving pixel coordinates $(u,v)$, $\tau \in \Re^4$,
%and the image gradients $\f_x, \f_y \in \Re^M$, the latter of which only has to be computed
%once per test image.  
The per-class alignment algorithm via ALM is summarized in Algorithm
\ref{alg:iterative_alignment}. 
%In terms of computational complexity, one significant
%term is the computation of the matrix pseudo-inverse $B_j^\dagger$, whose
%computational cost is bounded by $O(n_i^2m + n_i^3)$. However, for per-class face
%alignment, since both $B_j$ and $J_{j}$ are very tall matrices (\ie, $m\gg n_i$
%for all $i=1, \cdots, K$ classes), we find that the computation cost is still
%dominated by the frequently computed matrix-vector multiplications in the
%inner loop.
\begin{algorithm}[ht!]
\caption{\bf (Face Alignment via ALM)} \label{alg:iterative_alignment}
\small
{\bf Input:} $\bb$, $A_i$, $\x_0 = \mathbf{0}$, $\tau_0$, and $J_0$.
\begin{algorithmic}[1]
\WHILE{not converged ($j = 1,2,\ldots$)}
\STATE Update $\bb_j \leftarrow \frac{\bb\circ \tau_{j-1}}{\|\bb\circ \tau_{j-1}\|}$; $B_j= [A_i, -J_{j-1}]$ and corresponding $(B_j^\dagger)^T$
\STATE Initialize $\ww_0 = \mathbf 0$, $\yy_0 = \mathbf 0$
\WHILE{not converged ($k = 1,2,\ldots$)}
\STATE $\uu_0\leftarrow \ww_{k-1}$; $\zz_0\leftarrow \ee_{k-1}$
\WHILE{not converged ($l = 1,2,\ldots$)}
\STATE $\zz_l \leftarrow \mbox{shrink}\left(\bb_j - B_j\uu_{l-1} + \frac{\yy_{k}}{\mu_{k-1}}, \frac{1}{\mu_{k-1}}\right)$
\STATE $\uu_l \leftarrow B_j^\dagger \left(\bb_j - \zz_{l} + \frac{\yy_{k-1}}{\mu_{k-1}} \right) $
\ENDWHILE
\STATE $\ww_k \leftarrow \uu_l$; $\ee_k \leftarrow \zz_l$
\STATE $\yy_{k} \leftarrow \yy_{k-1} + \mu_{k-1} (\bb_j - B_j\ww_{k} - \e_{k})$
\STATE $\mu_{k} \leftarrow \rho\mu_{k-1}$
\ENDWHILE
\STATE Update $\ee_j$, $\tau_j$, and $J_j$
\ENDWHILE
\end{algorithmic}
{\bf Output:} $\tau_i^*\leftarrow \tau_j, \e_i^*\leftarrow \ee_j$
\end{algorithm}

@@

For the E5530, the L2 cache in each core is only able to store about 16 images.
For image alignment problem, the number of training images per class is
typically larger than 16.  
In this paper we compare two mappings of the parallelism available in the alignment
stage to the concurrency available on the CPU: a naive implementation that is purely 
library-based, and the optimized version we advocate.

The naive implementation disregards 
problem-level parallelism, and instead maps pixel-level parallelism
onto both the core-level and the SSE-level concurrency provided by the CPU.  In
other words, alignment is performed of the query
image against the training images of a single user at a time using all cores on
both processors.  The potential advantage of this implementation is that all of the variables
for the inner loop, most notably $B_j$ and $B_j^\dagger$, fit in L2 cache.

The optimized implementation maps problem-level parallelism onto core-level
concurrency, and maps pixel-level parallelism onto SSE-level concurrency.  In
other words, eight instances of \eqref{eq:alignment-linearization} are executed
in parallel with each problem running on one of the eight cores.  The advantage
of this implementation is that since the cores are operating on different
alignment problem instances, no data is shared between cores, and therefore no
synchronization is necessary between the cores.  Furthermore, even with eight
problem instances running concurrently, the local variables for the inner
loop still fit in L3 cache.  Because of the sequential data access patterns of
the solver, the CPU hardware is able to use the full L3 cache bandwidth, which is 
high enough to make the program CPU limited, rather than memory limited.  
For these reasons, we expect this implementation to out-perform
the previous implementation.

@@

\subsection{GPU Implementation} \vspace{-0.05in}
\label{sec:alignment_implementation_gpu}
While on the CPU, algorithm performance is highly dependent on effective
cache usage, on the GPU cache is relatively unimportant for ALM based
$\ell_1$-min.  As can be seen in Figure \ref{fig:caches}, the GPU has 
a very small amount of cache compared to the CPU.  While it might be possible
to fit a single instance of the alignment problem (with a slightly reduced problem size)
into L2 cache, this would not be an efficient use of the GPU's resources.
The strength of the GPU's memory architecture for our purposes is it's ability to sustain a very
high bandwidth to DRAM.  This bandwidth is achieved by having a very large number of
threads issuing interleaved memory requests.  Solving many alignment problem instances
concurrently increases the number of threads that can be used.

Therefore, our recommended GPU implementation is strikingly similar in spirit
to our recommended CPU implementation: on the CPU we use the cores to solve
multiple instances concurrently, and on the GPU we use multiple SMP's for the 
same purpose.
\footnote{In CUDA terms, our proposed alignment stage
implementation consists of a single kernel that performs alignment for all
subject classes.  Each instance of the alignment problem is assigned its own
thread block, and the GPU hardware schedules as many thread blocks to run
concurrently as the hardware will allow.} The number of subject classes that
are actually scheduled to run concurrently is chosen by the GPU hardware, 
but can be indirectly influenced by manual tuning of the code.\footnote{In CUDA terms, it is the kernel launch
configuration that is being tuned.} We have empirically determined
that performance is highest with 5-7 subject classes aligned simultaneously on each SMP.
%This mapping of the problem-level parallelism onto the SMP-level
%concurrency is illustrated in Figure \ref{fig:alignment_mapping_gpu}.
%\begin{figure}
%\centering
%\includegraphics[width=3.4in]{figures/alignment_mapping_gpu}
%\caption{Proposed mapping of alignment parallelism onto GPU concurrency}
%\label{fig:alignment_mapping_gpu}
%\end{figure}

Several other aspects of the implementation merit discussion.  
First of all, we take advantage of the GPU's special hardware dedicated to bilinear interpolation
for the computation of $b(\tau)$ and $J(\tau)$, which essentially consist of resampled
versions of the test image and its derivatives.
Second of all, since there are no standard CUDA libraries that work at the SMP level, we use a
custom routine for computing $B_j^\dagger = (B_j^TB_j)^{-1} B^T = G^{-1} B_j^T$. $G$
is inverted using Gaussian elimination with partial pivoting.  To achieve
precision comparable to the single precision LAPACK routines in Intel's MKL
with this simplistic algorithm, we use double precision.  Since $G$ is only $n
\times n$ and $B_j^\dagger$ is only computed once per optimization problem, the
cost of the inversion is dwarfed by the cost of other steps.  
Similarly, reductions (i.e. sum or $\ell_1$ norm) on image-sized vectors are also
performed in double precision.
%The lack of a atomic floating point addition on the GPU
%significantly reduces the perfomance of this step of the algorithm.
%In Section \ref{sec:experiment}, we show the performance of our CPU and GPU
%implementations running on synthetic and real image data.

>>>>

%\title{Accelerating $\ell_1$-Minimization Using Many-Core CPUs/GPUs \\ and Application to Face Recognition
%\title{Efficient Parallelization of Sparse Representation for Face Recognition}
%\thanks{Corresponding author: . This work was partially supported by ARO MURI W911NF-06-1-0076.}}
%\title{Many-Core Parallelization of $\ell_1$-Minimization for Face Recognition
\title{\vspace{-0.2in} Parallelization of Fast $\ell_1$-Minimization for Face Recognition}
%\title{Parallelization of Fast $\ell_1$-Minimization Towards Real-Time Face Recognition}

@@

\begin{abstract} Recently a family of promising face recognition algorithms
based on sparse representation and $\ell_1$-minimization ($\ell_1$-min) have been
developed.  These algorithms have not yet seen commercial application, 
largely due to higher computational cost compared to other traditional
algorithms. This paper studies techniques for leveraging the 
massive parallelism available in GPU and CPU hardware to accelerate
$\ell_1$-min based on augmented Lagrangian method (ALM) solvers. 
For very large problems, the GPU is faster due to higher memory bandwidth, while 
for problems that fit in the larger CPU L3 cache, the CPU is faster.
On both architectures, the proposed implementations significantly outperform
naive library-based implementations, as well as previous systems. 
The source code of the
algorithms will be made available for peer evaluation.  \end{abstract}
\vspace{-0.1in}

@@

In this paper, we study parallelization of $\ell$-min on many-core CPU and GPU
architectures. Although $\ell_1$-min \eqref{eq:l1min} is a convex
program, conventional algorithms such as interior-point methods
\cite{ChenS2001-SIAM,TibshiraniR1996} scale poorly
for applications such as face recognition. Recently, a number of
accelerated algorithms have been proposed that explicitly take advantage of
the special structure of $\ell_1$-min problems
\cite{LorisI2009,YangA2010-ICIP}. We investigate parallelization of a
state-of-the-art $\ell_1$-min solution based on
\emph{augmented Lagrangian methods} (ALM) \cite{BertsekasD2003,YangA2010-ICIP}.

@@

Although ALM has been applied to improve the speed of a prototype face recognition system \cite{WagnerA2011-PAMI},
due to the high per-class cost
of the alignment step, the system still fails to achieve \emph{real-time}
performance against datasets of hundreds or thousands of subjects.  In this paper,
in addition to accelerating the generic
$\ell_1$-min objectives \eqref{eq:l1min} and \eqref{eq:l1min_denoise}, we 
also discuss how to efficiently accelerate the face alignment step
\eqref{eq:l1min_alignment} on multi-core CPUs and GPUs.

@@

Finally, we present extensive benchmark to compare the performance of the
generic ALM $\ell_1$-min algorithm and the corresponding face recognition
routines on massively parallel CPUs and GPUs.  To this end, a face
recognition system for security and access control applications has been implemented, 
and is diagrammed in Figure \ref{fig:pipeline}. This pipeline follows the work of \cite{WagnerA2011-PAMI}, but
the fast implementations presented in this paper are novel.
Since the parallelization of any algorithm
is highly dependent on the target hardware, we will give a brief overview of
the architecture of our test system in detail in Section
\ref{sec:concurrency}.
Finally, note that 
while this paper focuses on the parallelism available in a single (shared memory) machine,
applications targeting thousands of subjects or more will likely require computation to be
distributed over a cluster or network of computers. These systems are not addressed here.
The source code of our implementations, which target NVIDIA Fermi GPUs
as well as Intel x86-64 CPUs, will be made available for peer evaluation.
\begin{figure}
\centering
{\tiny \input{figures/pipeline_simplified.tex}}
\caption{The face recognition pipeline.\vspace{-0.1in}}
\label{fig:pipeline}
\end{figure}

@@

One of the drawbacks of most interior-point methods for $\ell_1$-min is that they require the solution sequence to follow an
interior path via gradient descent or conjugate gradient methods, which are computationally expensive.
To mitigate these issues, an approach called \emph{Homotopy} has been recently studied to accelerate the
speed of $\ell_1$-min \cite{OsborneM2000,EfronB2004,MalioutovD2005,DonohoD2006}.
Although Homotopy can be shown to exactly estimate BP when the solution is
sufficiently sparse \cite{DonohoD2006}, the algorithm still involves
computationally expensive operations such as matrix-matrix multiplication and
linear least-squares problems with varying $A$
matrices.

%Homotopy methods for recovering sparse signals were
%first studied in the context of LASSO \cite{OsborneM2000}, which inspired a
%solution to the \emph{forward stagewise linear regression} problem called LARS
%\cite{EfronB2004}, and eventually led to the corresponding Homotopy algorithms
%for BP in \cite{MalioutovD2005,DonohoD2006}.

%\footnote{During the face alignment algorithm presented later, The $A$ matrix changes between alignment iterations but not within the L1 solver}  
In Section \ref{sec:ALM}, we contend that ALM is a better choice for
implementation on many-core CPUs and GPUs. The ALM algorithm belongs to a
category of approximate $\ell_1$-min solutions called \emph{iterative
shrinkage-thresholding} (IST) methods \cite{WrightS2008,BeckA2009}.  IST
algorithms mainly involve elementary operations such as vector algebra and
matrix-vector multiplication. Therefore, when the dimension of the problem
becomes high, IST-type algorithms are particularly suitable for hardware
systems with a high degree of concurrency. In \cite{YangA2010-ICIP}, the
authors showed that ALM is able to significantly improve the solver speed,
while achieving estimation accuracy competitive with other $\ell_1$-min
solutions. Therefore, we choose ALM as the core algorithm for
implementation of $\ell_1$-min in the parallel face recognition pipeline.

@@

In practice, the optimal values for the triplet $(\xx^*, \yy^*, \mu)$ are all
unknown. Furthermore, it has been observed that solving
\eqref{eq:optimal-ALM} with a large initial value of $\mu$ tends to lead to
slower convergence speed \cite{WrightS2008,YangA2010-ICIP}. In
\cite{BertsekasD2003,YangJ2009}, an alternating procedure has been shown to
iteratively update $\xx$ and $\yy$:
\begin{equation}
\left \{
\begin{array}{lll}
\xx_{k+1} & = & \arg\min_{\xx} \, L_{\mu_k} (\xx,\yy_k)\\
\yy_{k+1} & = & \yy_k + \mu_k (\bb - A\xx_{k+1}) \\
\end{array}
\right . ,
\label{eq:alm}
\end{equation}
where $\mu_{k}\rightarrow \infty$ increases monotonically.
The iteration terminates when the estimates $(\xx_k, \yy_k)$ converge.

@@

\subsection{GPU Hardware Concurrency} \vspace{-0.05in}
The main defining characteristics of contemporary multi-core GPU architectures
are that they have two (much wider) levels of concurrency, relatively small amounts of
 cache, and relatively low clock speeds.  Whereas most of the transistors on
a typical CPU are dedicated either to cache or to hardware that enables higher
clock speeds (such as branch prediction, out-of-order instruction execution, etc...),
most of the transistors on a GPU are dedicated to arithmetic logic units.
For our GPU implementations, we target NVIDIA Fermi GPU architecture (\eg, the GTX 480
GPU used in this paper).  An explanation of the GPU programming model (CUDA) at a useful level of completeness would take more space than is available here, so we will instead frame our discussion in terms of 
hardware capabilities. 
%\footnote{A discussion of the CUDA programming model can be found in the NVIDIA C for CUDA Programming Guide}
CUDA programmable GPUs are comprised of several
\emph{streaming multiprocessors} (SMP), each of which is roughly analogous to a CPU core.
For Fermi architecture GPU's, there are up to 16 SMP's, and each SMP is capable of executing up to 64 single precision 
floating point operations
concurrently. \footnote{In CUDA terms, the warp width is 32, and the floating pipeline can
issue up to two warps simultaneously}

@@

While the programming model for
the SIMD units of each SMP is somewhat more flexible than on the CPU,
leveraging the flexibility typically comes at the cost of reduced concurrency.\footnote{
In CUDA terms, code branches that cause warp divergence usually
result in serialization of the different groups of threads}  Therefore, for
the purposes of comparing hardware architectures, the SIMD units on the GPU are
roughly analogous to the CPU SIMD units. Thus, in summary, the GPU hardware 
provides two levels of concurrency: {\em SMP-level}
concurrency and {\em thread-level} concurrency.  Note that the GPU provides more
concurrency than a CPU at both levels (14 SMP's vs. 8 cores) and (64 wide vs. 4 wide SIMD
units).

@@

Face alignment \eqref{eq:l1min_alignment} estimates an image transformation
$\tau$ that rectifies the query image $\tilde{\bb}$ with possible pose
variation w.r.t. each training class $A_i$, which leads to the minimal sparsity
in error $\ee$ after the alignment. Note that directly solving
\eqref{eq:l1min_alignment} is inefficient since it is a non-convex problem
and may exhibit local minima.
However, given a good initial guess of the
transformation $\tau$ (\eg, provided by an accurate face detector), the optimal
solution for $\tau$ can be sought iteratively by linearization at each $j$th
step:
\begin{equation}
\min_{\xx, \ee, \Delta \tau_j}\|\ee\|_1\quad \subj\quad \tilde\bb\circ\tau_j +  J_j\Delta \tau_j = A_i\xx + \ee,
\end{equation}
where $J_j = \nabla_{\tau_j}(\tilde\bb\circ\tau_j)$ is the Jacobian, and
$\Delta \tau$ is a step update to $\tau$. Denote $\bb_j =
\tilde\bb\circ\tau_j$, $B_j = [A_i, -J_j]$ and $\ww^T = [\xx^T, \Delta
\tau^T]$, then the update $\Delta \tau$ can be computed by solving the
following linear program:
\begin{equation}
\min_{\ww, \ee}\|\ee\|_1\quad \subj\quad \bb_j = B_j\ww + \ee.
\label{eq:alignment-linearization}
\end{equation}
%If the warping $\tau$ is constrained to be a similarity transform,
%it can be paramaterized in such a way that the computation of $\frac{\partial \bb}{\partial \tau_p}$
%reduces to several linear vector operations involving pixel coordinates $(u,v)$, $\tau \in \Re^4$,
%and the image gradients $\f_x, \f_y \in \Re^M$, the latter of which only has to be computed
%once per test image.  
The per-class alignment algorithm via ALM is summarized in Algorithm
\ref{alg:iterative_alignment}. 
%In terms of computational complexity, one significant
%term is the computation of the matrix pseudo-inverse $B_j^\dagger$, whose
%computational cost is bounded by $O(n_i^2m + n_i^3)$. However, for per-class face
%alignment, since both $B_j$ and $J_{j}$ are very tall matrices (\ie, $m\gg n_i$
%for all $i=1, \cdots, K$ classes), we find that the computation cost is still
%dominated by the frequently computed matrix-vector multiplications in the
%inner loop.
\begin{algorithm}[ht!]
\caption{\bf (Face Alignment via ALM)} \label{alg:iterative_alignment}
\small
{\bf Input:} $\bb$, $A_i$, $\x_0 = \mathbf{0}$, $\tau_0$, and $J_0$.
\begin{algorithmic}[1]
\WHILE{not converged ($j = 1,2,\ldots$)}
\STATE Update $\bb_j \leftarrow \frac{\bb\circ \tau_{j-1}}{\|\bb\circ \tau_{j-1}\|}$; $B_j= [A_i, -J_{j-1}]$ and corresponding $(B_j^\dagger)^T$
\STATE Initialize $\ww_0 = \mathbf 0$, $\yy_0 = \mathbf 0$
\WHILE{not converged ($k = 1,2,\ldots$)}
\STATE $\uu_0\leftarrow \ww_{k-1}$; $\zz_0\leftarrow \ee_{k-1}$
\WHILE{not converged ($l = 1,2,\ldots$)}
\STATE $\zz_l \leftarrow \mbox{shrink}\left(\bb_j - B_j\uu_{l-1} + \frac{\yy_{k}}{\mu_{k-1}}, \frac{1}{\mu_{k-1}}\right)$
\STATE $\uu_l \leftarrow B_j^\dagger \left(\bb_j - \zz_{l} + \frac{\yy_{k-1}}{\mu_{k-1}} \right) $
\ENDWHILE
\STATE $\ww_k \leftarrow \uu_l$; $\ee_k \leftarrow \zz_l$
\STATE $\yy_{k} \leftarrow \yy_{k-1} + \mu_{k-1} (\bb_j - B_j\ww_{k} - \e_{k})$
\STATE $\mu_{k} \leftarrow \rho\mu_{k-1}$
\ENDWHILE
\STATE Update $\ee_j$, $\tau_j$, and $J_j$
\ENDWHILE
\end{algorithmic}
{\bf Output:} $\tau_i^*\leftarrow \tau_j, \e_i^*\leftarrow \ee_j$
\end{algorithm}

@@

For the E5530, the L2 cache in each core is only able to store about 16 images.
For image alignment problem, the number of training images per class is
typically larger than 16.  
In this paper, we compare two mappings of the parallelism available in the alignment
stage to the concurrency available on the CPU: a naive implementation that is purely 
based on stock libraries and compilers, and a manually optimized version we advocate.

The naive implementation disregards 
problem-level parallelism, and instead maps pixel-level parallelism
onto both the core-level and the SSE-level concurrency provided by the CPU.  In
other words, alignment of the query image is performed
against a single subject at a time using all available cores on
the processors.  The potential advantage of this implementation is that all of the variables
for the inner loop, most notably $B_j$ and $B_j^\dagger$, fit in L2 cache.

In contrast, our manually optimized implementation maps problem-level parallelism onto core-level
concurrency, and maps pixel-level parallelism onto SSE-level concurrency.  In
other words, eight instances of \eqref{eq:alignment-linearization} are executed
in parallel with each problem running on one of the eight cores.  The advantage
of this implementation is that since the cores are operating on different
alignment problem instances, no data is shared between cores, and therefore no
synchronization is necessary between the cores.  Furthermore, even with eight
problem instances running concurrently, the local variables for the inner
loop still fit in L3 cache.  Because of the sequential data access patterns of
the solver, the CPU hardware is able to use the full L3 cache bandwidth, which is 
high enough to make the program CPU limited, rather than memory limited.  
For these reasons, this implementation outperforms the previous naive solution.

@@

\subsection{GPU Implementation} \vspace{-0.05in}
\label{sec:alignment_implementation_gpu}
While on the CPU, algorithm performance is highly dependent on effective
cache usage, cache is relatively unimportant on the GPU for ALM based
$\ell_1$-min.  As can be seen in Figure \ref{fig:caches}, the GPU has 
a very small amount of cache compared to the CPU.  While it might be possible
to fit a single instance of the alignment problem (with a slightly reduced problem size)
into L2 cache, this would not be an efficient use of the GPU's resources.
The strength of the GPU's memory architecture for our purposes is its ability to sustain a very
high bandwidth to DRAM.  This bandwidth is achieved by having a very large number of
threads issuing interleaved memory requests.  Solving many alignment problem instances
concurrently increases the number of threads that can be used.

Therefore, our recommended GPU implementation is strikingly similar in spirit
to our recommended CPU implementation: on the CPU we use the cores to solve
multiple instances concurrently, and on the GPU we use multiple SMP's for the 
same purpose.\footnote{In CUDA terms, our proposed alignment stage
implementation consists of a single kernel that performs alignment for all
subject classes.  Each instance of the alignment problem is assigned its own
thread block, and the GPU hardware schedules as many thread blocks to run
concurrently as the hardware will allow.} The number of subject classes that
are actually scheduled to run concurrently is chosen by the GPU hardware, 
but can be indirectly influenced by manual tuning of the code (i.e., the kernel launch
configuration in CUDA terms). We have empirically determined
that performance is highest with 5-7 subject classes aligned simultaneously on each SMP.
%This mapping of the problem-level parallelism onto the SMP-level
%concurrency is illustrated in Figure \ref{fig:alignment_mapping_gpu}.
%\begin{figure}
%\centering
%\includegraphics[width=3.4in]{figures/alignment_mapping_gpu}
%\caption{Proposed mapping of alignment parallelism onto GPU concurrency}
%\label{fig:alignment_mapping_gpu}
%\end{figure}

Several other aspects of the implementation merit discussion.  
First of all, we take advantage of the GPU's special hardware dedicated to bilinear interpolation
for the computation of $b(\tau)$ and $J(\tau)$, which essentially consist of resampled
versions of the test image and its derivatives.
Second, since there are no standard CUDA libraries that work at the SMP level, we use a
custom routine for computing $B_j^\dagger = (B_j^TB_j)^{-1} B^T = G^{-1} B_j^T$, with $G$
inverted using Gaussian elimination with partial pivoting.  To achieve
precision comparable to the single precision LAPACK routines in Intel's MKL
with this simplistic algorithm, we use double precision.  Since $G$ is only $n
\times n$ and $B_j^\dagger$ is only computed once per optimization problem, the
cost of the inversion is dwarfed by the cost of other steps.  
Similarly, sums and dot-products of large vectors are also
performed in double precision.
%The lack of a atomic floating point addition on the GPU
%significantly reduces the perfomance of this step of the algorithm.
%In Section \ref{sec:experiment}, we show the performance of our CPU and GPU
%implementations running on synthetic and real image data.

\section{Parallelization of the Face Recognition Stage} \vspace{-0.05in}
\label{sec:recognition}
After the face alignment stage is complete, the 20 subject classes with lowest
alignment error are selected for recognition, $\bb$ and $A_i$ are re-sampled to a common alignment
using $\tau_i$, and $A_i$ are concatenated into a new matrix $A$.
A sparse representation of $\bb$ w.r.t $A$ is then recovered
by solving a single $\ell_1$-min problem, as shown
in \eqref{eq:l1min_denoise}, using Algorithm \ref{alg:alm_rec}.  The
coefficients $\x$ are then used to compute error residuals that are used for
classification.  
%This recognition algorithm has
%been extensively studied in
%\cite{WrightJ2009-PAMI,YangA2010-ICIP,WagnerA2011-PAMI}.
In this stage of the algorithm, there is no problem-level
parallelism to exploit, so this section will discuss how to exploit pixel-level parallelism
on both CPU and GPU hardware.
Then in Section \ref{sec:experiment}, we will benchmark the
performance of the two architectures and further demonstrate the speed
gains achieved by our proposed implementations over previously published implementations.
